---
title: "Work_on_MAIT_cells_PBMC"
author: "Chiara Milin"
date: "2026-08-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

THIS SCRIPT IS THE CONTINUATION OF THE PROJECT, BUT WITH PBMC SAMPLES. IT HAS NOT BEEN FINISHED.

## R Markdown

PBMC, SAT, and VAT were multiplexed, sequenced together, and subsequently demultiplexed. The tissue of origin for each cell can be identified from the Tissue column in the metadata, which annotates cells as PBMC, SAT, or VAT.

```{r cars}
library(dplyr)
library(Seurat)
library(patchwork)

data=readRDS("SeuObjxx_2.rds")
View(data)
data_filtered = data$Sample

seurat_pbmc = subset(data, subset = grepl("PBMC", Tissue, ignore.case = TRUE))
table(seurat_pbmc$Sample)
```

```{r}
metadata_patient = data.frame(patient_id=c('11183','11051','10113-1','10202','10380','10634','10738','9680','10205','9999','10113-2','9961','10742','10203','10291-2','11327','9932', '9991','11040','11471', '10291-1'), diabetes = c('No','No','No','No','No','No','No','No','No','No','No','No','No','Yes', 'Yes', 'Yes','Yes','Yes', 'Yes','Yes', 'Yes'), fibrosis = c('Fibrosis', 'Fibrosis', 'Fibrosis', 'Fibrosis', 'Fibrosis', 'Fibrosis', 'Fibrosis', 'Fibrosis', 'Fibrosis', 'No_fibrosis', 'No_fibrosis', 'No_fibrosis', 'Healthy', 'Fibrosis', 'Fibrosis', 'Fibrosis', 'Fibrosis', 'Fibrosis', 'Fibrosis', 'Fibrosis', 'Fibrosis'), fb_stage = c('F3', 'F1', 'F1', 'F1', 'F1', 'F1', 'F1', 'F1', 'F2', 'F0', 'F0', 'F0', 'H', 'F1', 'F1', 'F1', 'F2', 'F2', 'F3', 'F3', 'F1'))

library(tibble)
# 1. Temporary data.frame of metadata
meta_cellules = seurat_pbmc@meta.data %>%
  select(Patient_ID) %>%
  rownames_to_column("Cell_ID") %>%       # Saving the cells' names
  left_join(metadata_patient, by = c("Patient_ID" = "patient_id")) %>% 
  column_to_rownames("Cell_ID")           # Put them back in row names
seurat_pbmc = AddMetaData(object = seurat_pbmc, metadata = meta_cellules)
# Verification
head(seurat_pbmc@meta.data)
```

## Quality Control (QC) and selecting cells for further analysis

A few QC metrics commonly used include the number of unique genes detected in each cell, the total number of molecules detected within a cell and the percentage of reads that map to the mitochondrial genome (calculated by the function PercentageFeatureSet()).

```{r}
seurat_pbmc[["percent.mt"]] = PercentageFeatureSet(seurat_pbmc, pattern = "^MT-")
VlnPlot(seurat_pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

```{r}
seurat_pbmc = subset(seurat_pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
```

## Normalising the data

```{r}
seurat_pbmc = NormalizeData(seurat_pbmc, normalization.method = "LogNormalize", scale.factor = 10000)
```

# Identification of highly variable features

```{r}
seurat_pbmc = FindVariableFeatures(seurat_pbmc, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 = head(VariableFeatures(seurat_pbmc), 10)

# Plot variable features with and without labels
plot1 = VariableFeaturePlot(seurat_pbmc)
plot2 = LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```

# Scaling the data

```{r}
all.genes = rownames(seurat_pbmc)
seurat_pbmc = ScaleData(seurat_pbmc, features = all.genes)
```

# Perform linear dimensional reduction (PCA)

```{r}
seurat_pbmc = RunPCA(seurat_pbmc, features = VariableFeatures(object = seurat_pbmc), reduction.name = "pca") # I added reduction.name because PCA was not kept in the seurat package

# Examine and visualise PCA results in a few different ways
print(seurat_pbmc[["pca"]], dims = 1:5, nfeatures = 5)
```

```{r}
VizDimLoadings(seurat_pbmc, dims = 1:2, reduction = "pca")
```

```{r}
DimPlot(seurat_pbmc, reduction = "pca") + NoLegend()
```

```{r}
# Supervised approach
DimHeatmap(seurat_pbmc, dims = 1, cells = 500, balanced = TRUE)
```

# Determine the "dimensionality" of the dataset

```{r}
ElbowPlot(seurat_pbmc)
```

I am keeping 20 components.

# Cluster the cells

```{r}
seurat_pbmc = FindNeighbors(seurat_pbmc, dims = 1:20)
seurat_pbmc = FindClusters(seurat_pbmc, resolution = 0.3) # 16 communities
```

# Run non-linear dimensional reduction (uMAP/t-SNE)

```{r}
seurat_pbmc = RunUMAP(seurat_pbmc, dims = 1:20)
```

```{r}
DimPlot(seurat_pbmc, reduction = "umap")
```

# Finding differentially-expressed features (cluster biomarkers)

```{r}
seurat_pbmc = JoinLayers(seurat_pbmc)
# seurat_pbmc.markers = FindAllMarkers(seurat_pbmc, only.pos = TRUE)
# seurat_pbmc.markers %>%
  # group_by(cluster) %>%
  # dplyr::filter(avg_log2FC > 1)
# write.xlsx(seurat_pbmc.markers, "Marqueurs_par_cluster.xlsx", rowNames = FALSE) # All the markers in each cluster
```

- min.pct = 0.25 means that the gene must be expressed in at least 25 per cent of the cells in the cluster or comparison group

- logfc.threshold = 0.25 means that only genes with a log2FC > 0.25 are selected

- only.pos is not specified. By default: only.pos = FALSE

So I end up with:the genes that are overexpressed in the cluster (positive log2FC) and the genes that are underexpressed in the cluster (negative log2FC).

```{r}
seurat_pbmc.markers = FindAllMarkers(seurat_pbmc, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
top20_markers = seurat_pbmc.markers %>%
    group_by(cluster) %>%
    slice_max(order_by = avg_log2FC, n = 20, with_ties = FALSE) %>%
    ungroup()
library(openxlsx)
write.xlsx(top20_markers, "Final_selection_of_clusters_PBMC.xlsx", rowNames = FALSE)
DoHeatmap(seurat_pbmc, features = top20_markers$gene) + NoLegend()
```

I chose this line to select my top 20 markers because I want to know which genes are: enriched, specific or highly expressed, not which genes are absent.

Now I am going to examine the file "Final_selected_markers_PBMC.xlsx" to do research on the 20 most important markers for each cluster and try to annotate them.

Cluster 0 : CD4⁺ memory T cells, probably polarised Th2 (IL7R⁺ GATA3⁺ ICOS⁺ CD40LG⁺ CD28+ MAL+ LTB+ TNFRSF25+)

Cluster 1 : CD8⁺ cytotoxic effector memory T cells (ZNF683⁺ KLRG1⁺ CCL5⁺ GZMH⁺ CD8A/B+ GZMK+ LAG3+ KLRK1+) maybe exhausted

Cluster 2 : CD14⁺ inflammatory monocytes (VCAN⁺ IL1B⁺ NLRP3⁺ S100A12⁺ LYZ+ S100A9+)

Cluster 3 : NK-like cytotoxic γδ T cells (NCR1⁺ KLRF1⁺ PRF1⁺ GZMB⁺ TRDC+ TRGC1+ TRGC2+ CD3E+ KIR3DL1+ CD160+ IL2RB+ NCAM1+)

Cluster 4 : Naive T cells/T memory stem-like (CCR7⁺ TCF7⁺ LEF1⁺ MAL+ IL6ST+)

Cluster 5 : CD8 Naive / Central Memory T cells (CCR7⁺ LEF1⁺ NT5E⁺=CD73 CD8B+ CCR7+ LRRN3+)

Cluster 6 : Class-switched memory B cells (IGHA1⁺ IGHG2⁺ JCHAIN+ TNFRSF13B+=TACI FCRLA+ POU2AF1+)

Cluster 7 : Naive B cells (TCL1A⁺ FCER2⁺=CD23 IGHD⁺ IGHM⁺ CD19+ CD22+ PAX5+)

Cluster 8 : Quiescent T cells/long-lived memory cells (TOX+ BCL2+ MLLT3+ PRKCA+ PRKCH+ INPP4B+ ANK3+)

Cluster 9 : Non-classical/intermediate monocytes (CSF1R+ SIGLEC10+ MS4A7+ BATF3+ HMOX1+)

Cluster 10 : MAIT cells (TRAV1-2+ TRDV2+ TRGV9+ KLRB1+ SLC4A10+ ZBTB16+ IL18RAP+)

Cluster 11 : Erythrocyte contamination + adipocyte/endothelial markers (HBA1+ HBA2+ HBB+ FABP4+ SELENOP+ SPARCL1+)

Cluster 12 : Activated neutrophils (FCGR3B+ CXCR2+ IL1R2+ CXCL8+ MME=CD10+ ALPL+)

Cluster 13 : Basophils/Circulating mastocytes (MS4A2+ HDC+ CPA3+ FCER1A+ CCR3+ GATA2+)

CLuster 14 : Platelets / megakaryocytes (PPBP+ PF4+ GP9+ GP1BB+ TREML1+ ITGB3+)

Cluster 15 : Plasmacytoid dendritic cells (pDC) (LILRA4+ CLEC4C=BDCA2+ DNASE1L3+ CUX2+)


# Finding MAIT cells

Cluster 10 is already highly suspected to be MAIT cells, but I find it safer to create a score according to characteristical genes of MAIT cells.

```{r}
mait_genes = c("KLRB1", "SLC4A10", "CXCR6", "TRAV1-2", "ZBTB16")
mait_genes %in% rownames(seurat_pbmc) # check if they are present in the object
```

The idea is to calculate a MAIT-cell score for each cells, select the ones with the highest score and finally check the cluster they are in.

```{r}
seurat_pbmc_2 = AddModuleScore(seurat_pbmc, features = list(mait_genes), name = "MAIT_score")
```

Let's visualise where the richest cells in MAIT genes are.

```{r}
FeaturePlot(seurat_pbmc_2, features = "MAIT_score1", reduction = "umap")
```

Now, I identify the 5% richest cells in MAIT genes.

```{r}
threshold = quantile(seurat_pbmc_2$MAIT_score1, probs = 0.95)
mait_cells = WhichCells(seurat_pbmc_2, expression = MAIT_score1 > threshold)
```

Check which cluster they are in.

```{r}
table(seurat_pbmc_2$seurat_clusters[mait_cells])
```

This result suggests that cluster 3 has the highest MAIT cells signature, followed by clusters 10 and 0. Th others have low scores. However, the score may be influenced by activated T cells, so I will try a stricter method.

```{r}
exp = FetchData(seurat_pbmc_2, vars = mait_genes)
seurat_pbmc_2$MAIT_mean = rowMeans(exp)
quantile(seurat_pbmc_2$MAIT_mean, probs = c(0.5, 0.75, 0.9, 0.95, 0.99)) # check the scores to choose the threshold in the next line
mait_cells_2 = WhichCells(seurat_pbmc_2, expression = MAIT_mean > 1.2151541)

MAIT_cluster = subset(seurat_pbmc_2, cells = mait_cells_2)
VlnPlot(MAIT_cluster, features = mait_genes, group.by = "Patient_ID")
table(seurat_pbmc_2$seurat_clusters[mait_cells_2])
```

With this method, the result change radically and highlights that cluster 10 is by far the cluster containing the more MAIT cells.

```{r}
stages = c("F0", "F1", "F2", "F3")
comparisons = combn(stages, 2, simplify = FALSE)
results = lapply(comparisons, function(x) {
  FindMarkers(MAIT_cluster, ident.1 = x[1], ident.2 = x[2],group.by = "fb_stage")
})
names(results) = sapply(comparisons, function(x) paste(x, collapse = "_vs_"))
names(results)
```

# Combining F2 and F3 and H and F0

Thus, the study will be through fibrosis, not through MASLD.

```{r}
# Copying MAIT object
MAIT_F23 = MAIT_cluster
MAIT_F23$fibrosis_group = case_when(MAIT_F23$fb_stage %in% c("F2", "F3") ~ "F2-3", TRUE ~ as.character(MAIT_F23$fb_stage))
MAIT_F23$fibrosis_group = factor(MAIT_F23$fibrosis_group, levels = c("H", "F0", "F1", "F2-3"))
table(MAIT_F23$fb_stage)
table(MAIT_F23$fibrosis_group)

MAIT_F23 = MAIT_cluster
MAIT_F23$fibrosis_group = case_when(MAIT_F23$fb_stage %in% c("H", "F0") ~ "H_F0", MAIT_F23$fb_stage == "F1" ~ "F1", MAIT_F23$fb_stage %in% c("F2", "F3") ~ "F2-3")
MAIT_F23$fibrosis_group = factor(MAIT_F23$fibrosis_group, levels = c("H_F0", "F1", "F2-3"))
table(MAIT_F23$fibrosis_group)
```

To check if I could merge F2 and F3 and H and F0, I performed a differential analysis, finding the markers differentially expressed between the two groups in each pair. I found only 8 significantly differentially expressed genes out of 7,095 between F2 and F3 and 0 significantly differentially expressed genes between H and F0 (although it may be a bit more due to the low effective of 3 patients).

```{r}
# Pseudobulk DESeq2
counts_bis = GetAssayData(MAIT_F23, layer = "counts")
meta_bis = MAIT_F23@meta.data
library(Matrix)
pb_bis = rowsum(t(as.matrix(counts_bis)), group = meta_bis$Patient_ID)
pb_bis = t(pb_bis)
sample_metadata = meta_bis %>%
  select(Patient_ID, fibrosis_group) %>%
  distinct()
rownames(sample_metadata) = sample_metadata$Patient_ID
sample_metadata = sample_metadata[colnames(pb_bis),]
table(sample_metadata$fibrosis_group)
library(DESeq2)
dds_merged = DESeqDataSetFromMatrix(countData = pb_bis, colData = sample_metadata, design = ~ fibrosis_group)
dds_merged = DESeq(dds_merged)
```

```{r}
# Continuous trend analysis
meta_bis$fibrosis_score = case_when(meta_bis$fibrosis_group == "H"    ~ 0, meta_bis$fibrosis_group == "F0"   ~ 1, meta_bis$fibrosis_group == "F1"   ~ 2, meta_bis$fibrosis_group == "F2-3" ~ 3)

sample_metadata_merged = meta_bis %>%
  select(Patient_ID, fibrosis_score) %>%
  distinct()
rownames(sample_metadata_merged) = sample_metadata_merged$Patient_ID
sample_metadata_merged = sample_metadata_merged[colnames(pb_bis), ]

dds_merged2 = DESeqDataSetFromMatrix(countData = pb_bis, colData = sample_metadata_merged, design = ~ fibrosis_score)
dds_merged2 = DESeq(dds_merged2)
res_merged2 = results(dds_merged2)

gene_list = res_merged2$stat
names(gene_list) = rownames(res_merged2)
gene_list = gene_list[!is.na(gene_list)]
gene_list = sort(gene_list, decreasing = TRUE)

library(msigdbr)
library(fgsea)
hallmark_df = msigdbr(species = "Homo sapiens", category = "H")
pathways2 = split(hallmark_df$gene_symbol, hallmark_df$gs_name)

changing_pathways = fgsea(pathways = pathways2, stats = gene_list)
fgsea_filtered = changing_pathways %>%
  filter(padj < 0.05) # Keeping the most significant according to the adjusted p value
```

```{r}
fgsea_up = fgsea_filtered %>%
  filter(NES > 0) %>%
  arrange(desc(NES)) %>% # Positive NES ==> increase with disease
  select(pathway, NES, padj)
fgsea_up
```

```{r}
# Negative NES ==> decrease with disease
fgsea_down = fgsea_filtered %>%
  filter(NES < 0) %>%
  arrange(NES) %>%
  select(pathway, NES, padj)
fgsea_down
```

```{r}
# Visualization
fgsea_sig = fgsea_filtered %>%
  arrange(NES)
ggplot(fgsea_sig, aes(x = reorder(pathway, NES), y = NES, fill = NES)) + geom_col() + coord_flip() + scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) + theme_bw() + labs(x = "", y = "Normalized Enrichment Score (NES)")
```

```{r}
leading_edge_table_merged = fgsea_filtered %>%
  select(pathway, NES, leadingEdge) %>%
  unnest(leadingEdge)
res_merged2_df = as.data.frame(res_merged2) %>%
  tibble::rownames_to_column("leadingEdge")
leading_edge_counts_direction = leading_edge_table_merged %>%
  left_join(res_merged2_df %>%
  dplyr::select(leadingEdge, log2FoldChange), by = "leadingEdge") %>%
  dplyr::mutate(direction = ifelse(log2FoldChange > 0, "Up", "Down")) %>%
  dplyr::group_by(pathway, direction) %>%
  dplyr::summarise(n_genes = dplyr::n(), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = direction, values_from = n_genes, values_fill = 0)
leading_edge_counts_direction
```

# Exhaustion, activation and exhaustion scores for MAIT cells

```{r}
MAIT_F23 = AddModuleScore(MAIT_F23, features = list(cytotoxicity_genes), name = "Cytotoxicity")
MAIT_F23 = AddModuleScore(MAIT_F23, features = list(activation_genes), name = "Activation")
MAIT_F23 = AddModuleScore(MAIT_F23, features = list(exhaustion_genes), name = "Exhaustion")
VlnPlot(MAIT_F23, features = c("Cytotoxicity1", "Activation1", "Exhaustion1"), group.by = "fibrosis_group", pt.size = 0.2, ncol = 3)
```

```{r}
# It is far more rigorous to consider that a patient is a biological unit rather than cells
patient_scores_2 = MAIT_F23@meta.data %>%
  group_by(Patient_ID, fibrosis_group) %>%
  summarise(Cytotoxicity = median(Cytotoxicity1), Activation = median(Activation1), Exhaustion = median(Exhaustion1), .groups = "drop") %>%
  mutate(fibrosis_score = case_when(fibrosis_group == "H"    ~ 0, fibrosis_group == "F0"   ~ 1, fibrosis_group == "F1"   ~ 2, fibrosis_group == "F2-3" ~ 3))

cor.test(patient_scores_2$Cytotoxicity, patient_scores_2$fibrosis_score, method = "spearman")
cor.test(patient_scores_2$Activation, patient_scores_2$fibrosis_score, method = "spearman")
cor.test(patient_scores_2$Exhaustion, patient_scores_2$fibrosis_score, method = "spearman")
```

```{r}
cor.test(patient_scores_2$Cytotoxicity, patient_scores_2$Exhaustion, method = "spearman")
cor.test(patient_scores_2$Cytotoxicity, patient_scores_2$Activation, method = "spearman")
cor.test(patient_scores_2$Activation, patient_scores_2$Exhaustion, method = "spearman")
ggplot(patient_scores_2, aes(Cytotoxicity, Activation, color = fibrosis_group)) +
  geom_point(size = 4) + geom_smooth(method = "lm", se = FALSE)
```

```{r}
graphique_2 = patient_scores_2 %>%
  pivot_longer(cols = c(Exhaustion, Activation, Cytotoxicity), names_to = "Score_type", values_to = "Score")
graphique_2$stage_num = c("H" = 0, "F0" = 1, "F1" = 2, "F2-3" = 3)[graphique_2$fibrosis_group]
ggplot(graphique_2, aes(stage_num, Score)) + geom_point(size = 2) + geom_smooth(method = "lm", color = "gold", se = TRUE) + facet_wrap(~ Score_type, scales = "free_y") + theme_bw()
```

# Sub-clustering of MAIT cells

```{r}
library(Seurat)
MAIT = FindVariableFeatures(MAIT_cluster, selection.method = "vst", nfeatures = 2000)
head(VariableFeatures(MAIT))
MAIT = ScaleData(MAIT)
MAIT = RunPCA(MAIT, features = VariableFeatures(MAIT))
ElbowPlot(MAIT) # I am keeping 15 dimensions because the populations is already quite homogenous
MAIT = FindNeighbors(MAIT, dims = 1:15)
MAIT = FindClusters(MAIT, resolution = 0.2) # 6 communities
table(Idents(MAIT))
MAIT = RunUMAP(MAIT, dims = 1:15)
DimPlot(MAIT, label = TRUE)
markers_MAIT = FindAllMarkers(MAIT, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
markers_MAIT %>%
  group_by(cluster) %>%
  slice_max(avg_log2FC, n = 10)

```

```{r}
best_MAIT_markers = markers_MAIT %>%
  group_by(cluster) %>%
  slice_max(avg_log2FC, n = 20, with_ties = FALSE)
write.xlsx(best_MAIT_markers, "Best_MAIT_markers.xlsx", rowNames = FALSE)

VlnPlot(MAIT, features = c("TRAV1-2", "SLC4A10", "KLRB1", "CXCR6"), group.by = "seurat_clusters")
```

```{r}
cluster_prop_merged = MAIT_F23@meta.data %>%
  group_by(Patient_ID, fibrosis_group, seurat_clusters) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Patient_ID) %>%
  mutate(prop = n / sum(n))

cluster_summary = cluster_prop_merged %>%
  group_by(fibrosis_group, seurat_clusters) %>%
  summarise(mean_prop = mean(prop), se_prop = sd(prop) / sqrt(n()), .groups = "drop")

ggplot(cluster_summary, aes(x = fibrosis_group, y = mean_prop, colour = factor(seurat_clusters), group = seurat_clusters)) + geom_line(linewidth = 1) + geom_point(size = 3) + theme_classic() + labs(x = "Fibrosis stage", y = "Mean proportion per patient", colour = "MAIT cluster")
```

```{r}
library(dorothea)
library(viper)
library(dplyr)
data(dorothea_hs)

leading_edge_table_merged = fgsea_filtered %>%
  select(pathway, NES, leadingEdge) %>%
  unnest(leadingEdge)

# Only keep the interactions between TF and genes with high level of confidence
dorothea_highconf = dorothea_hs %>%
  dplyr::filter(confidence %in% c("A","B","C"))
# Associate each gene of the leading edge with his known transcription factors
pathway_tf = leading_edge_table_merged %>%
  inner_join(dorothea_highconf, by = c("leadingEdge" = "target")) # one gene can appear in several pathways and be regulated by several transcription factors
# Prepare the results of DESeq2
tf_expression = as.data.frame(res_merged2) %>%
  tibble::rownames_to_column("tf")
# Only look at transcription factors that are in the data
pathway_tf_expr = pathway_tf %>%
  inner_join(tf_expression, by = "tf")
# Link pathways and transcription factors
pathway_tf_summary = pathway_tf_expr %>%
  group_by(pathway, tf) %>%
  summarise(n_targets = n_distinct(leadingEdge), log2FC = dplyr::first(log2FoldChange), stat = dplyr::first(stat), pvalue = dplyr::first(pvalue), padj = dplyr::first(padj), .groups = "drop")
# Calculate a priority score
pathway_tf_summary = pathway_tf_summary %>%
  mutate(priority_score = abs(stat) * sqrt(n_targets))
# Classify the transcription factors
pathway_tf_summary = pathway_tf_summary %>%
  mutate(direction = case_when(stat > 0 ~ "Increases_with_fibrosis", stat < 0 ~ "Decreases_with_fibrosis", TRUE ~ NA_character_)) %>%
  arrange(desc(priority_score))
# Take out the best transcription factors for each pathway
top_tfs = pathway_tf_summary %>%
  group_by(pathway) %>%
  slice_max(order_by = priority_score, n = 10, with_ties = FALSE) # classify by priority score because transcription factors go to the nucleus and tend not to be detected really well in scRNA-seq so I don't give to much imortance to the significativity
# Export to Excel
library(openxlsx)
write.xlsx(top_tfs, file = "Top_TFs_pathways_F23_merged.xlsx", rowNames = FALSE)
```

The idea behind the score is to combine two pieces of information : the importance of the TF in MY study, represented by stat (the higher |stat| is, the more strongly the TF is associated with the progression of fibrosis) and the importance of the TF in the pathway, represented by n_targets (the more leading-edge genes a TF regulates, the more likely it is to contribute to the pathway signal). I use the square root of the number of targets in order to prevent genes with a really high number of targets being overestimated in their role in MASLD.


# Diabetes

```{r}
library(Matrix)
library(DESeq2)
pb = AggregateExpression(MAIT, group.by = "Patient_ID", assays = "RNA", return.seurat = FALSE)$RNA
colnames(pb) = sub("^g", "", colnames(pb))
meta_pb = unique(MAIT@meta.data[, c("Patient_ID", "diabetes")])
meta_pb = meta_pb[match(colnames(pb), meta_pb$Patient_ID),]

rownames(meta_pb) = meta_pb$Patient_ID
all(colnames(pb) == rownames(meta_pb))
meta_pb$diabetes = factor(meta_pb$diabetes, levels = c("No", "Yes"))
dds_diab = DESeqDataSetFromMatrix(countData = pb, colData = meta_pb, design = ~ diabetes)
dds_diab = DESeq(dds_diab)
res_diabetes = results(dds_diab)
```

```{r}
DimPlot(MAIT, group.by = "diabetes", reduction = "umap")
```

```{r}
# Proportion of each cluster according to the status
prop_df = MAIT@meta.data %>%
  group_by(diabetes, seurat_clusters) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(diabetes) %>%
  mutate(prop = n / sum(n))
ggplot(prop_df, aes(x = seurat_clusters, y = prop, fill = diabetes)) + geom_bar(stat = "identity", position = "dodge") + theme_classic()
```

Do diabetic patients have more or less MAIT cells than non diabetic patients ?

```{r}
# Counting per patient
mait_counts = MAIT@meta.data %>%
  group_by(Patient_ID, diabetes) %>%
  summarise(n_MAIT = n())
mait_counts
ggplot(mait_counts, aes(x = diabetes, y = n_MAIT, fill = diabetes)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.15, size = 3) + theme_classic() + labs(x = "Diabetes", y = "Number of MAIT cells")
```

```{r}
# Statistical analysis
wilcox.test(n_MAIT ~ diabetes, data = mait_counts) # no significant
```

```{r}
mait_counts %>%
  group_by(diabetes) %>%
  summarise(median = median(n_MAIT), mean = mean(n_MAIT), sd = sd(n_MAIT), n = n())
```

```{r}
# Total cells per patient
total_cells = table(seurat_liver$Patient_ID)
total_cells

# Total MAIT cells per patient
total_mait_cells = table(MAIT$Patient_ID)
total_mait_cells

# Recapitulative table
df_total = data.frame(patient = names(total_cells), total_cells = as.numeric(total_cells))
df_mait = data.frame(patient = names(total_mait_cells), mait_cells = as.numeric(total_mait_cells))
diabetes_info = seurat_liver@meta.data %>%
  select(Patient_ID, diabetes) %>%
  distinct()
proportion_table = df_total %>%
  left_join(df_mait, by = "patient") %>%
  mutate(mait_cells = replace_na(mait_cells, 0), percent_MAIT = (mait_cells / total_cells)*100)
proportion_table = proportion_table %>%
  left_join(diabetes_info, by = c("patient" = "Patient_ID"))
proportion_table
```

```{r}
ggplot(proportion_table, aes(x = diabetes, y = percent_MAIT, fill = diabetes)) + geom_boxplot(width = 0.6, alpha = 0.7, outlier.shape = NA) + geom_jitter(width = 0.15, size = 3, alpha = 0.8) + labs(x = "", y = "% of MAIT cells", title = "Proportion of MAIT cells according to diabetes status") + theme_classic() + theme(legend.position = "none")
```

```{r}
wilcox.test(percent_MAIT ~ diabetes, data = proportion_table)
```

```{r}
stage_info = seurat_liver@meta.data %>%
  select(Patient_ID, fb_stage) %>%
  distinct()
proportion_table = proportion_table %>%
  left_join(stage_info, by = c("patient" = "Patient_ID"))
proportion_table
```

# Number of MAIT cells according to the stade of MASLD

```{r}
stage_info = seurat_liver@meta.data %>%
  select(Patient_ID, fb_stage) %>%
  distinct()

proportion_table = proportion_table %>%
  left_join(stage_info, by = c("patient" = "Patient_ID"))
library(ggplot2)

ggplot(proportion_table, aes(x = fb_stage.x, y = mait_cells)) +
  geom_boxplot(outlier.shape = NA, fill = "lightblue") +
  geom_jitter(width = 0.15, size = 3) + theme_classic() +
  labs(x = "MASLD stage",
    y = "Number of MAIT cells",
    title = "Number of MAIT cells according to the disease stage")

ggplot(proportion_table, aes(x = fb_stage.x, y = percent_MAIT)) +
  geom_boxplot(outlier.shape = NA, fill = "salmon") +
  geom_jitter(width = 0.15, size = 3) + theme_classic() +
  labs(x = "MASLD stage",
    y = "% of MAIT cells",
    title = "Proportion of MAIT cells according to the MASLD stage")

summary_stage = proportion_table %>%
  group_by(fb_stage.x) %>%
  summarise(mean_MAIT = mean(mait_cells), sd_MAIT = sd(mait_cells), n = n())

ggplot(summary_stage, aes(x = fb_stage.x, y = mean_MAIT, group = 1)) +
  geom_line(size = 1.2) + geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_MAIT - sd_MAIT,
                    ymax = mean_MAIT + sd_MAIT), width = 0.1) + theme_classic() +
  labs(x = "MASLD stage", y = "Mean number of MAIT cells",
    title = "Evolution of MAIT cells across MASLD progression")
```


